---
title: "Farino Final Project"
output: html_document
---
## __Setup__
#### Load Seurat & tidyverse
#### Adjust memory limit to import large datasets
```{r}
library(Seurat)
library(tidyverse)
setwd("E:/R-data/Arvind")
memory.limit(size=20000)
Tsukui_Bleo <- readRDS("Tsukui_Bleo.rds")
```

## __Analyze__
#### Find Cluster that espresses GOI
```{r}
FeaturePlot(Tsukui_Bleo, "Acta2", cols = c("lightgreen", "darkblue"))
```

#### View clusters with labels
```{r warning=FALSE}
DimPlot(Tsukui_Bleo, label=TRUE, label.size=8) + NoLegend()
```

#### Import GPCR list from IUPHAR
```{r}
GPCRTargets <- read.csv("E:/R-data/Arvind/GPCRTargets.csv")
```

##### Pull out only the columnn that lists the GPCRs in the mouse gene format
```{r}
GPCRs <- GPCRTargets %>% select(MGI.symbol)
glimpse(GPCRs)%>% head(3)
```

#### Make a heatmap that compares GPCR expression between the clusters
```{r warning=FALSE}
DoHeatmap(Tsukui_Bleo, features = GPCRs [1:415,1], size=2) + NoLegend()
```

#### Featureplot of P2ry14 demonstrates the receptor is selectively expressed in fibroblasts
```{r}
FeaturePlot(Tsukui_Bleo, "P2ry14", cols = c("lightgreen", "darkblue"))
```
