---
title: "Tidybiology Final Project"
output:
  html_document: default
  pdf_document: default
---

Federica Mosti

## Loading Libraries

```{r message=FALSE}
library(tidyverse)
library(tidybiology)
library(ggrepel)
library(gridExtra)
library(readxl)
```

## Import Data 

First I imported the chromosome dataset from the tidybiology library. 
```{r}
data("chromosome")
glimpse(chromosome)
```

I was interested in human telomeres, so I created a dataset with data from public articles and  databases.

```{r include=FALSE}
Telomeres <- read_xlsx("/Users/federica/tidybiology/Telomeres.xlsx")
```
```{r}
glimpse(Telomeres)
```


## Prepare dataset 

I joint the dataset and prepared for labeling the graph
```{r}
new_chromosome <- left_join(chromosome, Telomeres, by= "id")

#I created a new column which will be useful later.
new_chromosome1 <- new_chromosome %>%
  mutate(last_gene=basepairs-Nearest_gene)

# I changed the order of the id column manually
new_chromosome2 <- new_chromosome1
new_chromosome2$id <- factor(new_chromosome2$id,                                   
                  levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11","12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "X", "Y"))
glimpse(new_chromosome2)
```
```{r include=FALSE}
Ychrom <- new_chromosome %>%
  arrange(telomere_arm_q_length_mbp_lymphocytes) %>%
  slice(1)
Secchrom <- new_chromosome %>%
  arrange(telomere_arm_q_length_mbp_lymphocytes) %>%
  slice(3)
```


## Exploratory Data Analysis
### Graph 1

I checked if there was a correlation between chromosome length and telomere length (of both arm p and q)  in lymphocytes.

```{r message=FALSE, fig.width= 9, fig.height= 7}
ggplot(new_chromosome) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes, color = "red"), size = 2) +
  geom_smooth(aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes), method = "lm", se = FALSE, col="red")+
  geom_text_repel(data=Ychrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes, label="Yq"))+
  geom_text_repel(data=Secchrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes, label="2q"))+
  geom_point(aes(x=length_mm, y = telomere_arm_p_length_mbp_lymphocytes, color = "blue"), size = 2)+
  geom_smooth(aes(x=length_mm, y = telomere_arm_p_length_mbp_lymphocytes), method = "lm", se = FALSE, col="blue")+
  geom_text_repel(data=Ychrom, aes(x=length_mm, y = telomere_arm_p_length_mbp_lymphocytes, label="Yp"))+
  geom_text_repel(data=Secchrom, aes(x=length_mm, y = telomere_arm_p_length_mbp_lymphocytes, label="2p"))+
  scale_colour_manual(name = 'Legend', guide = 'legend', values = c('red' = 'red', 'blue' = 'blue'),labels = c('Telomeres p arm','Telomeres q arm'))+
  labs(x = "Chromosome Lenght (mm)", y = "Telomeres length (mbp)", title = "Lenght of Human Telomeres in fuction of Chromosome Length", caption = "Data from Kelley et al., 2005")+
  scale_shape_manual(values = c("blue" = 1, "red" = 1))+
  theme_bw()
```

## Check other cell types
### Graph 2

Then, I wanted to check if the telomere of arm q of Y and 2nd chromosomes were shorter in other tissues, too.

```{r include=FALSE}
#Questo è il graphico per la legenda
mylegend <- ggplot(new_chromosome) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes, col = "red"), size = 2) +
  labs(x = "Chromosome Lenght", y = "Telomeres length")+
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_lung, col = "green"), size = 2) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_pancreas, col = "deepskyblue1"), size = 2) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_skin, col = "orange",), size = 2) +
  scale_color_manual(name = 'Legend', values = c("red", "green", "deepskyblue1", "orange"), labels = c("Lymphocytes", "Lungs", "Pancreas", "Skin"), guide = "legend")+
  theme(legend.position="right", legend.direction = "horizontal")
  theme_bw()
```

```{r fig.width= 8, fig.height= 6}
Lym <- ggplot(new_chromosome) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes), col = "red", size = 2) +
  geom_text_repel(data=Ychrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes, label="Yq"))+
  geom_text_repel(data=Secchrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_lymphocytes, label="2q"))+
  labs(x = "Chromosome Lenght (mm)", y = "Telomeres length (mbp)")+
  theme_bw()
Lung <-ggplot(new_chromosome) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_lung),col = "green", size = 2) +
  geom_text_repel(data=Ychrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_lung, label="Yq"))+
  geom_text_repel(data=Secchrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_lung, label="2q"))+
  labs(x = "Chromosome Lenght (mm)", y = "Telomeres length (mbp)")+
  theme_bw()
Pan <- ggplot(new_chromosome) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_pancreas), col = "deepskyblue1", size = 2) +
  geom_text_repel(data=Ychrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_pancreas, label="Yq"))+
  geom_text_repel(data=Secchrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_pancreas, label="2q"))+
  labs(x = "Chromosome Lenght (mm)", y = "Telomeres length (mbp)")+
  theme_bw()
  
Skin <-ggplot(new_chromosome) +
  geom_point(aes(x=length_mm, y = telomere_arm_q_length_mbp_skin), col = "orange", size = 2) +
  geom_text_repel(data=Ychrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_skin, label="Yq"))+
  geom_text_repel(data=Secchrom, aes(x=length_mm, y = telomere_arm_q_length_mbp_skin, label="2q"))+
  labs(x = "Chromosome Lenght (mm)", y = "Telomeres length (mbp)")+
  theme_bw()

extract_legend <- function(my_ggp) {
  step1 <- ggplot_gtable(ggplot_build(my_ggp))
  step2 <- which(sapply(step1$grobs, function(x) x$name) == "guide-box")
  step3 <- step1$grobs[[step2]]
  return(step3)
}
shared_legend <- extract_legend(mylegend)
grid.arrange(arrangeGrob(Lym, Lung, Pan, Skin, ncol = 2), shared_legend, nrow = 2, heights = c(10, 1), top = "Telomere length in different tissues")

```

## How many immunity-related genes are express in each human chromosome?
### Graph 3

I wanted to know if this shorter telomere in chromosome 2 depends on how many immunity-related genes were present in chromosome 2 and if that can lead to a more open chromatin and then, a shorter telomere.

```{r fig.width= 8, fig.height= 6}
ggplot(new_chromosome2, aes(x=id, y=Percentage_immune_system_genes)) +
  geom_segment(aes(x=id,xend=id, y=0, yend=Percentage_immune_system_genes), color=ifelse(new_chromosome2$id %in% c("2"), "red", "grey50"), size=ifelse(new_chromosome2$id %in% c("2"), 1.3, 0.7) ) +
  geom_point( color=ifelse(new_chromosome2$id %in% c("2"), "red", "grey50"), size=ifelse(new_chromosome2$id %in% c("2"), 5, 2) ) +
  labs(x = "Chromosome", y = "% of immune system genes", title = "Percentage of immunity-related genes in each chromosome", caption = "Data from innateDB")+
  theme(legend.position="none")+
  theme_bw()
```

## Are immunity-related genes in chromosome 2 nearer the telomeres?
### Graph 4

Finally, I wondered if the problem 

```{r fig.width= 8, fig.height= 6}
ggplot(new_chromosome2, aes(x=id, y=last_gene)) +
  geom_segment(aes(x=id,xend=id, y=0, yend=last_gene), color=ifelse(new_chromosome2$id %in% c("2"), "red", "grey50"), size=ifelse(new_chromosome2$id %in% c("2"), 1.3, 0.7) ) +
  geom_point( color=ifelse(new_chromosome2$id %in% c("2"), "red", "grey50"), size=ifelse(new_chromosome2$id %in% c("2"), 5, 2) ) +
  labs(x = "Chromosome", y = "Distance from the end of the chromosome (bp)", title = "Location of the gene nearest to the telomere", caption = "Data from innateDB")+
  theme(legend.position="none")+
  theme_bw()
```


## Conclusion

I actually don't knnow why chromosome 2 has a storter telomere in lymphocytes and all my hypothesis have proved to be wrong.

## Prioritized follow-up studies

* Check telomere length in lymphocytes with different techniques

* Check if in different kind of lymphocytes the telomeres are different

* Check epigenetic status of telomeres of chromosome 2 in lymphocytes 

* Check nuclear architecture

## Bibliography

Kelley J, de Bono B, Trowsdale J. IRIS: a database surveying known human immune system genes. Genomics. 2005 Apr;85(4):503-11. doi: 10.1016/j.ygeno.2005.01.009. PMID: 15780753.

Demanelis K, Jasmine F, Chen LS, Chernoff M, Tong L, Delgado D, Zhang C, Shinkle J, Sabarinathan M, Lin H, Ramirez E, Oliva M, Kim-Hellmuth S, Stranger BE, Lai TP, Aviv A, Ardlie KG, Aguet F, Ahsan H; GTEx Consortium, Doherty JA, Kibriya MG, Pierce BL. Determinants of telomere length across human tissues. Science. 2020 Sep 11;369(6509):eaaz6876. doi: 10.1126/science.aaz6876. PMID: 32913074.

https://www.innatedb.com/

## Acknoledgements

Shout out to Matt, Akshay for giving us some tools to actually code on R and my classmates for listening to me. 

### Session information for provenance and reproducibility

```{r}
utils:::print.sessionInfo(sessionInfo()[-8]) 
```

