---
title: "ABartlett_TidyBioPresentation_7AADEarlyInfection"
output: html_document
message: FALSE
warning: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

> EBV infection of Primary B Cells

This dataset was generated from infecting isolated B Cells with EBV strain B95.8. At the time of infection the cells were treated with 2 different drugs, either separately or together. Cell proliferation and cell death was evaluated via flow cytometry. 

```{r include=FALSE}
# install.packages("devtools")
# devtools::install_github("matthewhirschey/tidybiology")
library(tidybiology)
library(tidyverse)
library(paletteer)
ALI05_7AAD_Normalized <- read_csv("Data/ALI05_7AAD_Normalized.csv")
glimpse(ALI05_7AAD_Normalized)
```

## Proliferating B Cells
```{r, message=FALSE}
ALI05_7AAD_Normalized %>%
dplyr::filter(Sample == "Tx1284") %>%
filter(DPI %in% c(4,7,10,14)) %>%
ggplot(aes(DPI,Norm_BCellProlifs))+
  geom_smooth(aes(colour = Treatment))+
  theme_bw()+
  labs(x = "Days Post Infection", y = "Proliferating B Cells")+
  labs(title = "Donor Tx1284")
```

## Dead B Cells
```{r}
ALI05_7AAD_Normalized %>%
filter(Sample == "Tx1284") %>%
filter(DPI %in% c(4,7,10,14)) %>%
ggplot(aes(DPI,Norm_DeadBCells))+
  geom_point(aes(colour = Treatment))+
  theme_bw()+
  scale_color_paletteer_d("LaCroixColoR::Lime")+
  labs(x = "Days Post Infection", y = "Dead B Lymphocytes")+
  labs(title = "Donor Tx1284")
```

## Do dead proliferating B cells correlate with dead singlets?
```{r}
ALI05_7AAD_Normalized %>%
filter(Sample %in% c("Tx1284","Tx1285")) %>%
filter(DPI %in% c(4,7,10,14)) %>%
ggplot(aes(Norm_DeadSinglets,Norm_DeadBCellProlifs))+
  geom_point(aes(colour = Treatment))+
  theme_bw()+
  scale_color_paletteer_d("nord::aurora")+
  labs(x = "Dead Singlets", y = "Dead B Lymphocytes")+
  labs(title = "Donor Tx1284 and Tx1285")
```







