---
title: "Tidybiology Final Project"
author: "Luke Bartelt"
date: "`r format(Sys.time(), '%B %d, %Y')`"
---
## Libraries Required
```{r setup, message=FALSE}
library(tidyverse)
library(tidybiology)
```
```{r, include=FALSE}
#uncomment (or select) to clear environment; good for provenance
rm(list=ls()) 
```

### Data imported
First I imported the subcell dataset from the tidybiology library.

```{r import}
data("subcell")
```

### Subcell dataset
I observed the subcell dataset contained a 20,840 x 5 dataframe that lists all human proteins, their corresponding gene name, their Gene Ontology assignment, a score for subcellular speceficity, and what subcellular location they are located.

```{r glimpse, echo=FALSE}
glimpse(subcell)
```

## Exploratory Data Analysis (EDA.pt.1)
I then got interested in what chromosome these gene came from, so I also loaded the genes dataset. By using both of these datasets, I could link the subcellular localization of each gene to its chromosome of origin through the common gene_name column.

```{r}
data("genes")
glimpse(genes)
```

## EDA.pt.2  
In order to link the two datasets I first wanted to make sure the gene names were the same in each dataset, so I performed an intersect function on the gene_name column.

```{r}
# pipe into glimpse to get number of intersections, since intersect exceded max print length
intersect(genes$gene_name, subcell$gene_name) %>% glimpse()
```
This showed me that 18,688 genes would be retained after joining the two datasets.

## EDA.pt.3
Want to do an inner join function on gene_name so that my new dataset is not littered with "NAs" where gene_names differ. 
```{r}
genes.subcell <- inner_join(genes, subcell, by = "gene_name")
```

## EDA.pt.4
I then wanted to plot what proportion of genes from each chromosome were localized to a specific subcellular location to see if there were any interesting differences. For this a percent stacked bar chart made sense to me. I found a way to do this with ggplot2 within `geom_bar`.

```{r}
# load ggplot 
library(ggplot2)

# construct bar chart with position=fill aes setting
ggplot(genes.subcell, aes(fill='location', x='chromosome_scaffold_name', y='location')) +
  geom_bar(position="fill", stat='identity')
```

This failed horribly, because I did not have numerical values for each subcellular localization.

## EDA.pt.5
So I made a new dataframe that took advantage of the `group_by` and `summarise` functions to count observations by each chromosome. 

```{r}
location.counts <- genes.subcell %>% group_by(chromosome_scaffold_name, location) %>% summarise(n=n()) %>% as.data.frame()
location.counts %>% head()
```

## EDA.pt.6

chromosome_scaffold_name type is character so need to convert to numeric.

```{r, echo=FALSE, warning=FALSE}
location.counts$chromosome_scaffold_name <- as.numeric(location.counts$chromosome_scaffold_name)
# this gives NA for anything not directly numeric convertible

location.counts1 <- na.omit(location.counts)
#omit NAs - unfortunately removes X and Y :/

location.counts1 %>% head()
```

## Final Graph

Created a Stacked Bar Chart with 'geom_bar' setting 'aes(fill=location)' column to stack the bar chart by subcellular location and set 'position=fill' to make the bar proportional rather than vary by magnitude. I imported a custom color panel to make it shine.

```{r, include=FALSE}
library(RColorBrewer)
# Define the number of colors you want
nb.cols <- 15
mycolors <- colorRampPalette(brewer.pal(11, "Spectral"))(nb.cols)
```


```{r, echo=FALSE}
# construct bar chart with position=fill aes setting
ggplot(location.counts1, aes(fill=location, x=chromosome_scaffold_name, y=1)) +
  geom_bar(position="fill", stat='identity') +
  scale_x_continuous(breaks = 1:22) +
  labs(title = 'Proportion of Proteins by Subcellular Location and Chromosome', x = 'Chromosome', y = 'Proportion by Subcellular Location') +
  scale_fill_manual(values = mycolors)
```

## Conclusions
From this graph it seems that in general, each autosome contains a roughly equal percentage of genes to each subcellular location - with a few notable exceptions. Based on this data, chromosome 8 does not have any proteins localized to Vesicles, Lysosome, or Golgi. Chromosome 13, 14, 20, and 21 are also missing many subcellular locations.
  
## Prioritized follow-up studies
Ribosomal genes are missing from many chromosomes, which goes against prior observations in the field, which points out to me that this dataset is slighty incomplete. To follow up, I would do some extra datamining to find a more curated chromosomal gene position list. Many of the 'chromosome_scaffold_name' entries were not directly convertible to numeric, and prevented me from including them in my analysis, which may be skewing my results. If a more curated dataset held the same conclusion - I would want to do some comprehensive ribosomal microscopy experiments, knocking down genes as referenced in the databases to see if they held true.

## Acknolwedgements
Shout out to Matt, Akshay, and Allie for actually teaching the concepts of coding rather than just putting a textbook into powerpoint and reading it out loud. Big shout out to Madison who helped me get work out creative solutions to filter my data.

#Session information for provenance and reproducibility
```{r, include=FALSE}
#PROBABLY BEST TO LEAVE THIS HERE, BUT NO NEED TO PRINT IN THE FINAL REPORT.
utils:::print.sessionInfo(sessionInfo()[-8]) 
#You can remove an item from sessionInfo(), which is a list with a class attribute, by printing the resulting object omitting one of the list items (omitted list of packages installed, but not loaded)
```

