---
title: "Tidybiology Final Project"
author: "Faraz Ahmed Butt"
date: "`r format(Sys.time(), '%B %d, %Y')`" #put in quotes to make YAML recognize it
---
## Load libraries
```{r setup, include=FALSE}
library(tidyverse)
library(tidyr)
library(knitr)
library(tidybiology)
#uncomment (or select) to clear environment; good for provenance
#rm(list=ls()) 
```
`r library(tidyverse)`

## Import data
I loaded the subcell, protein and genes data from Tidybiology and combined them based on gene name.
```{r import, include=FALSE}
data(subcell)
data(proteins)
data(genes)
subcell_prot <- left_join(proteins,subcell, by="gene_name")
subcell_fin <- left_join(subcell_prot,genes,by="gene_name")
prot_loc <- select(subcell_fin,protein_name,location)
#IMPORT YOUR DATA
#MAKE SURE TO GET THE CODE CHUNK OPTIONS SET
```

## Take a `glimpse` into the data in the dataframe
The data is a combined data from the prtoeins, subcell and gene data frames available in Tidybiology. The data describes the proteins and their respective gene information along with their subcellular location.
```{r glimpse}
glimpse(subcell_fin)
```
## EXPLORATORY DATA ANALYSIS (EDA)
I began by first location at the abundance of proteins in each subcellular location. However I did not find something interesting.
```{r, echo=FALSE}
#COMMENCE ROLLING
#MAKE COMMENTS
#WRITE WHAT YOU SEE
ggplot(subcell_fin, aes(x = location, y = log(mass), color=location))+
 geom_jitter()+theme(axis.text.x = element_text(angle = 90))+labs(x = "Cellular Compartment", y="log(Protein Mass)",title="Complete Cell Proteome Size in Each Cellular Compartment")

```

## EDA Graph  
Then I thought to look for differences in a class of proteins, so I decided to look at the kinase distribution across cellular compartments and found something interested yet intuitive.  

```{r, echo=FALSE}
#FIRST I FAILED.
neww<-  filter(subcell_fin, grepl('kinase', protein_name))
ggplot(neww, aes(x = location, y = log(mass), color=location))+
  geom_jitter()+theme(axis.text.x = element_text(angle = 90))+labs(x = "Cellular Compartment", y="log(Protein Mass)",title="All Kinase Size in Each Cellular Compartment")
neww_fin<-select(neww,protein_name,location)
centrosome_1 <- filter(neww_fin,location=="Centrosome")
cytoplasm_1 <- filter(neww_fin,location=="Cytoplasm")
centrosome <- filter(prot_loc,location=="Centrosome")
cytoplasm<- filter(prot_loc,location=="Cytoplasm")
ER_1 <- filter(neww_fin,location=="ER")
Extracellular_1 <- filter(neww_fin,location=="Extracellular")
ER <- filter(prot_loc,location=="ER")
Extracellular<- filter(prot_loc,location=="Extracellular")
Membrane_1 <- filter(neww_fin,location=="Membrane")
Mitochondria_1 <- filter(neww_fin,location=="Mitochondria")
Membrane <- filter(prot_loc,location=="Membrane")
Mitochondria<- filter(prot_loc,location=="Mitochondria")
Nucleolus_1 <- filter(neww_fin,location=="Nucleolus")
Nucleus_1 <- filter(neww_fin,location=="Nucleus")
Nucleolus <- filter(prot_loc,location=="Nucleolus")
Nucleus<- filter(prot_loc,location=="Nucleus")
fin_data <- data.frame("location"=c("Centrosome","Cytoplasm","ER","Extracellular","Membrane","Mitochondria","Nucleolus","Nucleus"),"Ratio_Total_Kinase"=c(dim(centrosome_1)[1]/dim(centrosome)[1], dim(cytoplasm_1)[1]/dim(cytoplasm)[1],dim(ER_1)[1]/dim(ER)[1],dim(Extracellular_1)[1]/dim(Extracellular)[1],dim(Membrane_1)[1]/dim(Membrane)[1],dim(Mitochondria_1)[1]/dim(Mitochondria)[1],dim(Nucleolus_1)[1]/dim(Nucleolus)[1],dim(Nucleus_1)[1]/dim(Nucleus)[1]))
```
  
  **The cell membrane, cytoplasm and the nucleus clearly have a higher number of kinases then the other compartments.**  
  
## Final Graph

```{r final_graph, echo=FALSE}
#MAKE IT SHINE
ggplot(fin_data, aes(x = location, y = Ratio_Total_Kinase, color=location))+
  geom_jitter()+theme(axis.text.x = element_text(angle = 90))+ylim(0,0.1)+
  labs(x = "Cellular Compartment", y="Kinases/Total Proteins",title="Ratio of Kinases to Total Proteins in Each Cellular Compartment")

ggsave(here::here("6_final", "Kinases_Ratio.png"), plot = last_plot(), height = 5, width = 7, units = "in", dpi = 300)
```

## Conclusions
Even though the cell membrane, cytoplasm and nucleus have a higher number of kinases, however the ratio of kinases to total proteins in each subcellular compartment does not reflect that. With the centrosome have a much higher kinase to total protein ratio then the rest of the compartments.
  
## Prioritized follow-up studies
Look at signalling events that occur at centrosome during cell division, and how many of these are dependent on kinases.

## Acknolwedgements
I want to thank the instructors of the class.

#Session information for provenance and reproducibility
```{r, echo=FALSE}
#PROBABLY BEST TO LEAVE THIS HERE, BUT NO NEED TO PRINT IN THE FINAL REPORT.
utils:::print.sessionInfo(sessionInfo()[-8]) 
#You can remove an item from sessionInfo(), which is a list with a class attribute, by printing the resulting object omitting one of the list items (omitted list of packages installed, but not loaded)
```

